---
title: "Introdução ao R para análise exploratória de dados"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Apresentando o R

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Apresentando o R

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Apresentando o R

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Apresentando o R

* This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

* When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Apresentando o RStudio

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Apresentando o RStudio

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.


## Apresentando o RStudio

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Operações Básicas no R

```{r, echo = TRUE}
#Adição
2+2

#Subtração
2*2

#Divisão
2/2

#Potenciação
2^2
```

## Funções Básicas no R

```{r, echo = TRUE}
#Raiz quadrada
sqrt(4)
#Logaritmo na base 10
log(2,10)
#Logaritmo na base e
log(2)
#exponencial
exp(2)
#Fatorial
factorial(2)
```


## Funções Básicas no R

```{r, echo = TRUE}
#Valor absoluto
abs(-2)
#Arredondando
round(exp(2),2)
#função Piso
floor(log(2))
#função teto
ceiling(log(2))


```

## Tipo de Objetos: vetor, matriz, arrays, data-frames, listas e funções

* Vetor
```{r , echo = TRUE}
x <- c(3,1,5,4)
#Selecionando o segundo valor do vetor
x[2]
#Sequência
y <- 1:10
y
z <- seq(1,10,length = 10)
z
```


## Vetor


```{r , echo = TRUE}
#Repetindo o valor 6 quatro vezes
rep(6,4)
#Repetindo o vetor quatro vezes
rep(c(1,2,3),4)
#Repetindo o vetor quatro vezes
rep(c(1,2,3),each = 4)
#soma dos elementos de um vetor
sum(c(1,2,3))
```


## Vetor
```{r , echo = TRUE}
#Comprimento de um vetor
length(c(1,2,3))
#Ordenando os elementos de um vetor ... crescente
sort(c(2,1,3),decreasing = FALSE)
#Ordenando os elementos de um vetor ... decrescente
sort(c(2,1,3),decreasing = TRUE)
```


## Vetor
```{r , echo = TRUE}
#retorna a posição dos elementos ...
order(c(2,8,-4), decreasing = FALSE)
#retorna a posição dos elementos ...
order(c(2,8,-4), decreasing = TRUE)
#retorna a posição dos elementos ...
pmax(2,c(1,1,2,3))
#retorna a posição dos elementos ...
pmin(1,c(0,1,2,3))

```

## Matriz
```{r , echo = TRUE}
M <- matrix(c(1,2,3,4),ncol = 2,nrow = 2)
M
#Selecionando os valores
M[1,1]
M[,1]
M[2,]
```

## Matriz
```{r , echo = TRUE}
M <- matrix(0,ncol = 2,nrow = 2)
M
# ...  os valores
M[1,1] <- 1
M[2,] <- c(2,2)
M
```

## Matriz
```{r , echo = TRUE}
#Matriz diagonal
M <- diag(c(1,2))
M
# multiplicando matriz
M%*%M
```

## Matriz
```{r , echo = TRUE}
D <- matrix(c(1,2,3,4),ncol = 2,nrow = 2)
D
# matriz transposta
t(D)
# matriz inversa
solve(D)
```

## Funções

**Criando uma função:**

nome da função <- function(argumentos)$\{$ comandos da função  $\}$

**Exemplo:** Considere uma função que retorna a média de um vetor.

```{r , echo = TRUE}

z <- c(1,3,2,5,3,6)

zbar <- function(x){ sum(x)/length(x)}
zbar(z)


```


## Funções do R

**Operadores de comparação**

* Igualdade: ==
* Não igual: !=
* Menor: <
* Maior: >
* Menor ou igual: <=
* Maior ou igual: >=

**Operadores lógicos**

E: &&
Ou: ||

**Condicionais**

if (Expressão teste) $\{$
declaração
$\}$ else  declaraçãos

```{r , echo = TRUE}
x <- 3
if(x >= 1 & x <= 6){print("sim")} else {print("Não")}

```


## Funções do R - Ciclos

**While**

```{r , echo = TRUE}
x <- 3
while(x < 5){ x = x + 1 ; print(x)}

```
**For**
```{r , echo = TRUE}
x <- c()
for(i in 0:4){x[i+1] <- 2*i+1}

x
```

## Funções do R - Ciclos

replicate()
familia apply

falar sobre os pacotes ...


## Função Sample

```{r , echo = TRUE}
x <- c("Cara", "Coroa")
#lançando uma moeda
sample(x,5,replace = T)


```

## Distribuições de Probabiliade no R

* p for "probability", the cumulative distribution function (c. d. f.)
* q for "quantile", the inverse c. d. f.
* d for "density", the density function (p. f. or p. d. f.)
* r for "random", a random variable having the specified distribution

**Simulando variáveis aleatorias no R**

* $rnorm(n,mu,sigma)$ gerando $n$ valores da distribuição Normal($\mu, \sigma$). 
* $runif(n,a,b)$ gerando $n$ valores da distribuição Uniforme($a,b$).
* $rexp(n,lambda)$ gerando $n$ valores da distribuição Exponencial($\lambda$). 
* $rpois(n,lambda)$ gerando $n$ valores da distribuição Poison($\lambda$). 
* $rbinom(n,K,p)$ gerando $n$ valores da distribuição Binomial($K,p$). 
 




## Análise descritiva no R

```{r , echo = TRUE}

pesoH <- rnorm(100,80,5)
pesoZ <- rnorm(100,60,4)

summary(pesoH)

mean(pesoH)

sd(pesoH)

var(pesoH)

```

## Análise descritiva no R
tentar juntar tudo
```{r , echo = TRUE}
hist(pesoH)

```
## Análise descritiva no R box plot


## Gráficos ggplot2



## Ajustando um modelo de regressão no R










